---
title: "3.WGCNA_WT"
author: "Athmaja Viswanath"
date: "2024-09-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#0. LOADING REQUIRED PACKAGES
```{r loading packages}
library(WGCNA)
library(DESeq2)
library(tidyverse)
library(ggplot2)
library(CorLevelPlot)
library(gridExtra)

theme_set(theme_classic())
```

#1.Data preparation 
##A.Separating orthologous gene names so that they can be used to filter readcounts for each species.
Total number of orthologous genes detected is 13785 genes across C. remanei and C. latens (gene list from Daniel)
```{r}
# Load ortholog data from a tab-separated file, ignoring comments
orthologs = read.table("../1-Input/new_1to1_orthologgenelist.txt", sep="\t", header=TRUE, comment.char="#")

# Extract the relevant columns into separate variables
cre_ortho = orthologs[, 1]            # First column: cre_ortho, Cre names
clat_ortho = orthologs[, 2]           # Second column: clat_ortho, Clat names
cumulative_ortho = orthologs[, 3]     # Third column: cumulative_ortho, joint names

# Display the first few rows of the data frames
head(orthologs)
head(cre_ortho)
head(clat_ortho)

##Number of genes 
nrow(orthologs)
length(cre_ortho)
length(clat_ortho)
```

##B Filtering readcoutns for all samples to retain orthologous genes and combining to make one readcount file
Total number of genes in C. remanei = 20091
Total number of gene sin C. latens = 23073
Total number of 1:! orthologous genes = 13785

```{r}
#########
#C.remanei
#########
# Load counts for C. remanei
cre_counts = read.table("../1-Input/C.remanei_cumulative.txt", sep="\t", header=TRUE, row.names=1, comment.char="#")
# Look at the structure of data
head(cre_counts)
colnames(cre_counts)
nrow(cre_counts) #20091
# Filter to include only orthologs and exclude Whole animal samples
cre_counts_ortho = cre_counts[cre_ortho, -c(13:15)] 
# Update row names to match cumulative orthologs
rownames(cre_counts_ortho) = cumulative_ortho 
# Check the number of rows and column names
nrow(cre_counts_ortho)
colnames(cre_counts_ortho)

#########
#C.latens
#########
# Load counts for C. latens
clat_counts = read.table("../1-Input/C.latens_cumulative.txt", sep="\t", header=TRUE, row.names=1, comment.char="#")
# Look at the structure of data
head(clat_counts)
colnames(clat_counts)
nrow(clat_counts) #23073
# Filter to include only orthologs and exclude Whole animal samples
clat_counts_ortho = clat_counts[clat_ortho, -c(13:15)]
# Update row names to match cumulative orthologs
rownames(clat_counts_ortho) = cumulative_ortho
# Check the number of rows and column names for C. latens
nrow(clat_counts_ortho)
names(clat_counts_ortho)


# Combine ortholog counts from C. remanei and C. latens
wt_counts_orthologs = cbind(cre_counts_ortho, clat_counts_ortho)

# Display current column names
colnames(wt_counts_orthologs)

# Clean up column names by removing the "X" prefix
colnames(wt_counts_orthologs) = gsub("X", "", colnames(wt_counts_orthologs))
colnames(wt_counts_orthologs)
```

##C.Quality control of samples and outlier detection

```{r}
gsg = goodSamplesGenes(t(wt_counts_orthologs))
summary(gsg)

# Display counts of good genes and samples
table(gsg$goodGenes) #number of genes that are outliers = FALSE
table(gsg$goodSamples)

# Filter good genes and check dimensions
counts_orthologs = wt_counts_orthologs[gsg$goodGenes, ]
dim(counts_orthologs)

# Heirarchical clustering to detect outliers

htree = hclust(dist(t(counts_orthologs)), method = "average")
plot(htree) 


##PCA for detecting outliers
pca = prcomp(t(counts_orthologs))
pca_data = pca$x ####PCA calculations for all the samples

#Calculating the variance explained by each principal component

pca.var = pca$sdev^2
pca.var.percent = round(pca.var/sum(pca.var)*100, digits = 2)

pca_data = as.data.frame(pca_data)

ggplot(pca_data, aes(PC1, PC2)) +
  geom_point() +
  geom_text(label = rownames(pca_data)) +
  labs( x = paste0("PC1:", pca.var.percent[1], "%"), 
        y = paste0("PC2:", pca.var.percent[2], "%"))
```

#2. Normalization of data usign DESeq2
##A.Preparing coldata
```{r}
# Prepare the coldata data frame
sample_name = colnames(wt_counts_orthologs)  # Get sample names
tissue = substr(sample_name, 3, 3)            # Extract tissue type from sample names
sex = substr(sample_name, 2, 2)               # Extract sex from sample names
species = c(rep("Cre", 12), rep("Clat", 12))  # Assign species labels
batch = rep(c(1, 2, 3),8)                # Assign batch numbers

# Create the coldata data frame
coldata = data.frame(sample_name, species, sex, tissue, batch)

# Convert columns to factors for analysis
coldata$species = factor(coldata$species)
coldata$tissue = factor(coldata$tissue)
coldata$sex = factor(coldata$sex)
coldata$batch = factor(coldata$batch)

# Display the coldata data frame
coldata  # Note: WM samples are not included
```

```{r}
# Prepare column data by setting the first column as row names
coldata2 = coldata[, -1]  # Remove the first column
rownames(coldata2) = coldata[, 1]  # Set the first column as row names

# Check if all row names in coldata2 match the column names of counts_orthologs
all(rownames(coldata2) %in% colnames(counts_orthologs))
all(rownames(coldata2) == colnames(counts_orthologs))

# Create DESeqDataSet from count matrix and column data
dds = DESeqDataSetFromMatrix(
  countData = counts_orthologs, 
  colData = coldata2, 
  design = ~1  # Not specifying a model
)

# Remove low count genes (at least 15 counts in 12 samples)
dds75 = dds[rowSums(counts(dds) >= 15) >= 12, ]
nrow(dds75)#11935 genes 

# Check the number of genes retained
nrow(dds75)  # Should output 11935 genes

# Perform variance stabilization transformation
dds_norm = vst(dds75)

# Extract normalized counts for downstream analysis
norma.counts = assay(dds_norm) %>% t()

View(norma.counts)
```

#3. Network construction 
##A. Selection soft threshold and powers
```{r}
##Choose a set of soft threshold powers

power = c(c(1:10), seq(from = 12, to = 50, by = 2))

sft = pickSoftThreshold(norma.counts,
                        powerVector = power,
                        networkType = "signed",
                        verbose = 5)

sft.data = sft$fitIndices ##we will use the R suqared value and mean connectivity , max R^2 and min mean connectivity

###Visualization to pick the right power
a1 = ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, colour = "red") +
  labs(x = "Power", y = "Scale free topology model fit, signed R^2") +
  theme_classic()


a2 = ggplot(sft.data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  #geom_hline(yintercept = 0.8, colour = "red") +
  labs(x = "Power", y = "Mean Connectivity") +
  theme_classic()

grid.arrange(a1, a2, nrow = 2)  ##we need higher R^2 and low mean connectivity


###converting matric to numeric

norma.counts[] = sapply(norma.counts, as.numeric)  ##different step than official tutorial

soft_power = 18

temp_cor = cor #to prevent WGCNa from using other cor function

cor = WGCNA::cor

```

##B. Constructing the network

```{r}
###Memory estimate wrt blocksize

##can take 20-30 mins to run

bwnet = blockwiseModules(norma.counts, 
                         maxBlockSize = 14000, ##depends on the ram of the system 4gb = 8-10k, 16gb = 20,000, 232gb = 30,000 
                         TOMType = "signed", 
                         power = soft_power,
                         networkType = "signed",
                         mergeCutHeight = 0.25,#threshold that we want to merge similar modules at
                         numericLabels = FALSE, #want the module names to be colours if not, then say TRUE
                         randomSeed = 1234,
                         verbose = 3) 

cor = temp_cor

#2. Plotting dendrogram using network constructed

# Plot the dendrogram with color assignments for both unmerged and merged modules
plotDendroAndColors(bwnet$dendrograms[[1]], 
                    cbind(bwnet$unmergedColors, bwnet$colors), 
                    c("unmerged", "merged"), 
                    dendroLabels = FALSE,
                    oddGuide = TRUE,
                    hang = 0.03, 
                    guideHang = 0.05)


dendo1 = plotDendroAndColors(bwnet$dendrograms[[1]], 
                             cbind(bwnet$colors),
                             c("merged"),
                             dendroLabels = FALSE,
                             oddGuide = TRUE, 
                             hang = 0.03,
                             guideHang = 0.05)

```

##C. Saving module eigen gene information
```{r}
# Extract module eigengenes from the WGCNA network and view them
module_eigengenes = bwnet$MEs
#View(module_eigengenes)

# Save module eigengene values to a CSV file
write.csv(module_eigengenes, file = "../3-Output/module eigengenes values_WT.csv")

###CALCULATE AVERAGE ACROSS REPLICATES FOR EACH MODULE

# Count the number of genes in each module
table(bwnet$colors)

# Retrieve names of the modules
module_names = unique(bwnet$colors)

# Create a data frame for module colors and add row names as a new column
bwnet_df = as.data.frame(bwnet$colors)
bwnet_df = rownames_to_column(bwnet_df)
#View(bwnet_df)

# Save the data frame of module genes to a text file
write.table(bwnet_df, file = "../3-Output/WGCNA_modulegenes.txt",row.names = FALSE, col.names = FALSE, quote = FALSE)
```

#4. Identifying modules significantly associated with sex, tissue and species differences
##A. Binarizing trait data
```{r}
###1.Sexes - has three categories - M, F, W
#coldata2$sex = factor(coldata2$sex, levels = c("M", "F", "W")) #M will nto showup after binarised below

sex.out = binarizeCategoricalColumns(coldata2$sex,
                                     includePairwise = TRUE,
                                     includeLevelVsAll = FALSE)

row.names(sex.out) = row.names(coldata2) #need to change rownames when binarizing categorical variables


###2. Species
# species = coldata2 %>%
#   mutate(species_binary = ifelse(grepl("Cre", species), 1, 0)) %>% #merges M & W into 0s
#   select(5)

species.factor <- factor(coldata2$species,
                         levels = c('Clat','Cre','H1','H2')) #changing the reference level
species = binarizeCategoricalVariable(coldata2$species,
                                      includePairwise = TRUE,
                                      includeLevelVsAll = FALSE) ##o is the reference category
row.names(species) = row.names(coldata2) #need to change rownames when binarizing categorical variables

###3. TISSUES

tissues = binarizeCategoricalVariable(coldata2$tissue,
                                      includePairwise = TRUE,
                                      includeLevelVsAll = FALSE) ##o is the reference category
row.names(tissues) = row.names(coldata2) #need to change rownames when binarizing categorical variables


##Combining binarized data 

traits = cbind(sex.out, species, tissues)

##define number of genes and samples

nSample = nrow(norma.counts)
nGenes = ncol(norma.counts)
```
##B. Calculating corrrealtions between eingenes and traits

```{r}
# Calculate correlations between module eigengenes and various traits
modules.trait.correlation.sp = cor(module_eigengenes, species, use = "p") # Correlation with species
modules.trait.correlation.sex = cor(module_eigengenes, sex.out, use = "p") # Correlation with sex
modules.trait.correlation.tissue = cor(module_eigengenes, tissues, use = "p") # Correlation with tissues

# Overall correlation with all traits
modules.trait.correlation = cor(module_eigengenes, traits, use = "p") 

#Calculating p value for the correlations
modules.trait.corr.pvals.sp = corPvalueStudent(modules.trait.correlation.sp, nSample)
modules.trait.corr.pvals.sex = corPvalueStudent(modules.trait.correlation.sex, nSample)
modules.trait.corr.pvals.tissue = corPvalueStudent(modules.trait.correlation.tissue, nSample)

modules.trait.corr.pvals = corPvalueStudent(modules.trait.correlation, nSample)
```


##C. Extracting genes from each module
```{r}
# Map genes to modules
module.gene.mapping = bwnet_df %>% column_to_rownames(var = "V1")

# Define modules of interest
module_names
gene_lists = list()

# Loop through each module and extract genes
for (module in module_names) {
  gene_lists[[module]] = module.gene.mapping %>%
    filter(V2 == module) %>%
    rownames()
}

# Access genes using gene_lists$green or gene_lists$brown

```

#5. Calculating module membership

```{r}
module.mem.measure = cor(module_eigengenes, norma.counts, use = "p")

modules.mem.measure.pval = corPvalueStudent(module.mem.measure, nSample)

module.mem.measure = as.data.frame(module.mem.measure) #need to be transposed

modules.mem.measure.pval[1:5, 1:5] # depending on the p value , we can identify whihc genes have sig values for membership meaning can be hub genes
# 
# View(t(modules.mem.measure.pval))
# View(t(module.mem.measure))
```

#6. Identifying important genes that are different across "SPECIES"
##A. Calculating gene significance associated with difference in species
we use this measure, for yellow and black to make figures in the next section and 4.Figures_WGCNA_WT.Rmd
```{r}
# Calculate correlation between normalized counts and species for all genes
gene.sig.cor = cor(norma.counts, species, use = "p")
gene.sig.cor.pval = corPvalueStudent(gene.sig.cor, nSample)

# Convert correlation results to a data frame
gene.sig.cor = as.data.frame(gene.sig.cor)
#view(gene.sig.cor)

# Identify top genes significantly associated with the species trait
gs.sig.sp = gene.sig.cor.pval %>%
  as.data.frame() %>%
  arrange(Cre.vs.Clat) %>% # Sort by p-value (lowest to highest)
  filter(Cre.vs.Clat < 0.05) # Filter for significant genes

# Get the number of significant genes and total genes
num_sig_genes = nrow(gs.sig.sp)
total_genes = nrow(gene.sig.cor)
```

##B. Identifying intramodular hub genes in Yellow (M10) and Black (M4) modules
This is also plotted in 4A & 4B in the script 4.Figures_WGCNA_WT.Rmd
```{r}
####################
#YELLOW MODULE (M10)
####################

# Get genes in the yellow module
yellow_module = module.gene.mapping %>% 
  filter(V2 == "yellow") %>% 
  rownames()

# Get module membership values for genes in yellow module (M10)
# Note: module.mem.measure is transposed, so rows represent module names and columns represent gene names

mem_yellow = as.data.frame(t(module.mem.measure[2,yellow_module]))
gs_yellow = as.data.frame(abs(gene.sig.cor[yellow_module, 1]))

# Combine membership and gene significance data
yelow_gs_meme = cbind(mem_yellow, gs_yellow)

## Create a new column to categorize genes as "hub" or "non-hub"
yelow_gs_meme$category = with(yelow_gs_meme, ifelse(MEyellow>0.9 & `abs(gene.sig.cor[yellow_module, 1])`>0.9, "hub genes", "non-hub"))

####Number of hubgenes in yellwo/M10 module
table(yelow_gs_meme$category)

#Saving hub and non-hub gene lists
write.table(yelow_gs_meme, file = "../3-Output/yellow_gs_mem.txt", row.names = TRUE, col.names = TRUE, quote = FALSE)


#####Gene significance vs module membership plot
ggplot(yelow_gs_meme, aes(x=yelow_gs_meme$MEyellow, y=yelow_gs_meme$`abs(gene.sig.cor[yellow_module, 1])`, colour = yelow_gs_meme$category)) +
  geom_point(alpha = 0.6, show.legend = FALSE)+
  coord_cartesian(ylim = c(0, 1))+
  coord_cartesian(xlim = c(0, 1))+
  geom_hline(yintercept = 0.9, colour = "red")+
  geom_vline(xintercept = 0.9, colour = "red")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)))+
  ggtitle("Gene Significance vs Module membership for M10")


###################
# BLACK MODULE (M4)
###################

# Get genes in the black module
black_module = module.gene.mapping %>% 
  filter(V2 == "black") %>% 
  rownames()

# Get module membership values for genes in yellow module (M10)
# Note: module.mem.measure is transposed, so rows represent module names and columns represent gene names

mem_black = as.data.frame(t(module.mem.measure[8,black_module]))
gs_black = as.data.frame(abs(gene.sig.cor[black_module, 1]))

# Combine membership and gene significance data
black_gs_meme = cbind(mem_black, gs_black)

## Create a new column to categorize genes as "hub" or "non-hub"
black_gs_meme$category = with(black_gs_meme, ifelse(MEblack>0.9 & `abs(gene.sig.cor[black_module, 1])`>0.9, "hub genes", "non-hub"))

####Number of hubgenes in yellwo/M10 module
table(black_gs_meme$category)

#Saving hub and non-hub gene lists
write.table(black_gs_meme, file = "../3-Output/black_gs_mem.txt", row.names = TRUE, col.names = TRUE, quote = FALSE)

#####Gene significance vs module membership plot
ggplot(black_gs_meme, aes(x=black_gs_meme$MEblack, y=black_gs_meme$`abs(gene.sig.cor[black_module, 1])`, colour = black_gs_meme$category)) +
  geom_point(alpha = 0.6, show.legend = FALSE)+
  coord_cartesian(ylim = c(0, 1))+
  coord_cartesian(xlim = c(0, 1))+
  geom_hline(yintercept = 0.9, colour = "red")+
  geom_vline(xintercept = 0.9, colour = "red")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)))+
  ggtitle("Gene Significance vs Module membership for M4")

```

##C. Getting genelists for species-biased hub-genes for gprofiler
###I. Getting hub genes from Yellow (M10 module)
```{r}
# Read gene names from text files
remanei_gene_names = read.table("../1-Input/remanei_gene_names.txt", fill = T)
latens_gene_names = read.table("../1-Input/latens_gene_names.txt", fill = T)

# Check the structure of yellow module data
str(yelow_gs_meme)

# Extract hub genes and separate their names
yellow_genes = subset(yelow_gs_meme, yelow_gs_meme$category=="hub genes") %>% 
  rownames_to_column() %>% 
  separate(rowname, sep = 23, into = c("Crenames", "Clatnames"))

# Further separate Clatnames and remove the random column
yellow_genes = separate(yellow_genes, Clatnames, sep = 1, into = c("random", "Clatnames"))[,-2]

# Get corresponding names from remanei and latens gene lists
yellow_genes_cre = subset(remanei_gene_names, remanei_gene_names$V4 %in% yellow_genes$Crenames)[,2]
yellow_genes_clat = subset(latens_gene_names, latens_gene_names$V4 %in% yellow_genes$Clatnames)[,2]

# Get the number of yellow genes in the cre dataset
length(yellow_genes_cre)
length(yellow_genes_clat)

# Uncomment to write the results to text files
# write.table(yellow_genes_cre, file = "../3-Output/M10_yellow_crehubgenes.txt",
#             row.names = FALSE, col.names = FALSE, quote = FALSE)
# 
# write.table(yellow_genes_clat, file = "../3-Output/M10_yellow_clathubgenes.txt",
#             row.names = FALSE, col.names = FALSE, quote = FALSE)
```

###II. Getting hub genes from Black (M4 module)
```{r}
# Read gene names from text files
remanei_gene_names = read.table("../1-Input/remanei_gene_names.txt", fill = T)
latens_gene_names = read.table("../1-Input/latens_gene_names.txt", fill = T)

# Check the structure of yellow module data
str(black_gs_meme)

# Extract hub genes and separate their names
black_genes = subset(black_gs_meme, black_gs_meme$category=="hub genes") %>% 
  rownames_to_column() %>% 
  separate(rowname, sep = 23, into = c("Crenames", "Clatnames"))

# Further separate Clatnames and remove the random column
black_genes = separate(black_genes, Clatnames, sep = 1, into = c("random", "Clatnames"))[,-2]

# Get corresponding names from remanei and latens gene lists
black_genes_cre = subset(remanei_gene_names, remanei_gene_names$V4 %in% black_genes$Crenames)[,2]
black_genes_clat = subset(latens_gene_names, latens_gene_names$V4 %in% black_genes$Clatnames)[,2]

# Get the number of black genes in the cre dataset
length(black_genes_cre)
length(black_genes_clat)

# Uncomment to write the results to text files
# write.table(black_genes_cre, file = "../3-Output/M4_black_crehubgenes.txt",
#             row.names = FALSE, col.names = FALSE, quote = FALSE)
# 
# write.table(black_genes_clat, file = "../3-Output/M4_black_clathubgenes.txt",
#             row.names = FALSE, col.names = FALSE, quote = FALSE)
```

##D. Overlap between hub genes in Yellow (M10) and Black (M4) modules and species-biased genes (from DESeq2)

```{r}
crexclat_res <- read.csv("../3-Output/crexclat_genes.csv")

cre_biased = subset(crexclat_res, crexclat_res$crexclat == 1) %>% select("X")
clat_biased = subset(crexclat_res, crexclat_res$crexclat == -1) %>% select("X")

##Overlap with Yellow (M10) genes
M10_hubgenes = subset(yelow_gs_meme, yelow_gs_meme$category=="hub genes") %>% row.names()
M10_allgenes  = yelow_gs_meme %>% row.names()
M10_clat_overlap = clat_biased %>% filter(X %in% M10_hubgenes) #as M10 has Clat-biased genes only


##Overlap with Black (M4) genes
M4_hubgenes = subset(black_gs_meme, black_gs_meme$category=="hub genes") %>% row.names()
M4_allgenes  = black_gs_meme %>% row.names()
M4_cre_overlap = cre_biased %>% filter(X %in% M4_hubgenes) #as M4 has Cre-biased genes only





```


#7.Identifying important genes that are different across "SEXES"
##A. Calculating gene significance associated with difference in sex

```{r}
# Calculate correlation between normalized counts and sex for all genes

gene.sig.cor.sex = cor(norma.counts, sex.out, use = "p")
gene.sig.cor.pval.sex = corPvalueStudent(gene.sig.cor.sex, nSample)

# Convert correlation results to a data frame
gene.sig.cor.sex = as.data.frame(gene.sig.cor.sex)
view(gene.sig.cor.sex)

# Identify top genes significantly associated with the sex trait
gs.sig.sex = gene.sig.cor.pval.sex %>%
  as.data.frame() %>%
  arrange(data.M.vs.F) %>% ##from lowest to highest p-value
  filter(data.M.vs.F < 0.05)

# Get the number of significant genes and total genes
nrow(gs.sig.sex)
nrow(gene.sig.cor.sex)

```

##B. Identifying intramodular hub genes in sex-biased modules
##C. Getting genelists for overlapping sex-biased hub-genes for gprofiler
##D. Overlap between hub genes in modules and sex-biased genes (from DESeq2)
#8. Identifying important genes that are different across "TISSUES"
##A. Calculating gene significance associated with difference in tissues
```{r}
# Calculate correlation between normalized counts and tissues for all genes

gene.sig.cor.ts = cor(norma.counts, tissues, use = "p")
gene.sig.cor.pval.ts = corPvalueStudent(gene.sig.cor.ts, nSample)

# Convert correlation results to a data frame
gene.sig.cor.ts = as.data.frame(gene.sig.cor.ts)
view(gene.sig.cor.ts)

# Identify top genes significantly associated with the sex trait
gs.sig.ts = gene.sig.cor.pval.ts %>%
  as.data.frame() %>%
  arrange(S.vs.G) %>% ##from lowest to highest p-value
  filter(S.vs.G < 0.05)

# Get the number of significant genes and total genes
nrow(gs.sig.ts)
nrow(gene.sig.cor.ts)

```

##B. Identifying intramodular hub genes in tissue-biased modules
##C. Getting genelists for overlapping tissue-biased hub-genes for gprofiler
##D. Overlap between hub genes in modules and tissue-biased genes (from DESeq2)

